Resonant Scattering in a Uniform Sphere with Large Optical Depth

### 2021-07-18 ms.tex
Filling out gamma nm description

We seek now to calculate all eigenmodes $J_{nm}(\sigma)$ and eigenfrequencies $\gamma_{nm}$ for all spatial $n$ and for all eigenmodes $m$ at that given $n$. A resonant response in $J(n,\sigma,-i\gamma)$ is produced for special values of $\gamma \equiv \gamma_{nm}$. To measure the size of the response to detect where resonances occur, we sum the absolute value of $J(n,\sigma,-i\gamma)$ over $\sigma$ in array form. Call this response $f$, and use the index $j$ to represent the value of the response at discrete points over a range of $\gamma$. In places where $f_j > f_{j-1}$ and $f_j>f_{j+1}$, we have bounded a resonance that occurs in the interval $(\gamma_{j-1},\gamma_{j+1})$. We then call the ``solve'' step to hone in on the eigenvalue before continuing the sweep in $\gamma$. In the solve step, we evaluate the eigenfunctions at the points $(\gamma_{j-1},\gamma_{j},\gamma_{j+1})$ to obtain the responses $f_{j-1}$, $f_j$, and $f_{j+1}$. A ``guess'' at the correct eigenvalue $\gamma_{nm}$ can be calculated from
\be
\gamma_{\rm guess} &=& \frac{b\gamma_{j-1} - \gamma_{j+1}}{b - 1}
\ee
where
\be
b &=& \left(\frac{f_{j} - f_{j-1}}{f_{j} - f_{j+1}}\right)\left(\frac{\gamma_{j+1}-\gamma_{j}}{\gamma_{j-1}-\gamma_{j}}\right).
\ee
This is repeated by replacing former points $(\gamma_{j-1},\gamma_{j},\gamma_{j+1})$ with closer estimates while the size of the response grows as the resonance is approached. Given a sufficiently accurate eigenvalue $\gamma_{nm}$, we now find the corresponding eigenfunction $J_{nm}(\sigma)$. We can write down an approximate expression for the response near the resonances as
\be
J(n,\sigma,-i\gamma) & \simeq & \frac{J_{nm}(\sigma)}{\gamma_{nm}-\gamma} + C(\sigma),
\ee
where $C(\sigma)$ is weakly dependent on $\gamma$, and so we treat it as a constant. We can now solve for $J_{nm}(\sigma)$ by comparing two points near the resonance:
\be
J(n,\sigma,-i\gamma_1)  - J(n,\sigma,-i\gamma_2)& = &  \frac{J_{nm}(\sigma)}{\gamma_{nm}-\gamma_1} -  \frac{J_{nm}(\sigma)}{\gamma_{nm}-\gamma_2}
\ee
where $C(\sigma)$ has cancelled out in the difference. We can immediately solve as
\be
J_{nm}(\sigma) & \simeq & \frac{ J(n,\sigma,\gamma_1)  - J(n,\sigma,\gamma_2) }{ 1/(\gamma_{nm}-\gamma_1) - 1/(\gamma_{nm}-\gamma_2)}.
\ee
\begin{figure}
    \centering
    \includegraphics{Jsoln_n1_m100.pdf}
    \caption{Eigenfunctions $J_{nm}(x)$ for the lowest-order spatial eigenmode $n=1$, and a range of frequency eigenmodes up to $m=100$. Their oscillatory forms have varying amplitudes which sum to create the final form of the mean intensity. The largest contribution always comes from the ($n=1, m=1$) lowest-order eigenfunction.}
    \label{fig:jsoln}
\end{figure}
A set of eigenfunctions $J_{nm}(\sigma)$ are plotted in Fig. \ref{fig:jsoln} to illustrate their relative scales for different frequency eigenmodes $m$ at a fixed spatial eigenmode $n$. The overall scale of the $J_{nm}(\sigma)$ are set by the factor $E/(kR^3)$ with $E$ arbitrarily set to 1, $k = \tau_0 \sqrt{\pi} \Delta / R$, and $R = 10^{11}$ cm. With this normalization, an eigenfunction has typical size ${\sim} E a / \left(R^2 \Delta \right) = 10^{-37}$. Additional terms add smaller, faster-oscillating components that lead to higher accuracy upon summation with the lower-order terms. The oscillations of various modes must cancel perfectly to recreate the form of the solution in Eq. \ref{eq:Jrsigmat}, so many modes $m$ and $n$ are required for sufficient numerical precision in the final result. Too few (or imprecisely-calculated) modes creates ``ringing'' in the function that indicates the presence of imperfect cancellations in the positive and negative contributions of individual poles.

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{gamma_nm.pdf}
    \caption{The scaling of numerically-obtained resonant frequencies as compared with the analytic expression in Eq. \ref{eq:gamma_nm}.}
    \label{fig:gamma_nm}
\end{figure}

TODO: The values of the $\gamma_{nm}$ resonant frequencies are calculable analytically. Their scale depends on $m$, $n$, and the factor $(a\tau_0)^{1/3}$ according to 
\be \label{eq:gamma_nm}
2**(-1/3) * np.pi**(13/6)*n**(4/3)*(m-7/8)**(2/3)*fc.clight/p.radius/(p.a * p.tau0)**(1/3.) 
\ee
The size of additional eigenfrequencies scales as a powerlaw in $m$, yet as is seen in Fig. \ref{fig:gamma_nm} the scaling is weak, requiring up to 500 $m$ to reduce the scale of $\gamma_{nm}^{\ \ -1}$ by two orders of magnitude.

A single eigenmode $J_{nm}(\sigma)$ is oscillatory out to some turning point, $\sigma_{tp}$, at which point the function becomes evanescent. The location of the turning point can be found by ignoring the delta-function discontinuity at the source frequency $\sigma_s$ in Eq. \ref{eq:diffusion_plugged_in} and re-examining the resulting homogeneous differential equation. We obtain
\be
\frac{d^2J}{d\sigma^2} & = & \left[ \left( \frac{\kappa_n \Delta }{k} \right)^2 - \frac{3\phi \gamma\Delta^2}{ck}\right] J
\ee
where the line profile can be written approximately as $\phi = (2a/\pi)^{1/3}/(3 \Delta |\sigma|^{2/3})$. When the coefficient on the right hand side is positive, exponential growth or decay solutions are found (evanescence). This occurs on the line wings. When the coefficient on the right hand side is negative, oscillatory solutions are found (propagation), which occurs in the line core. The boundary between propagation and evanescence occurs at the turning point, given by
\be
\sigma_{tp} & = & \sqrt{\frac{2a}{\pi}}\left( \frac{k \gamma}{ \kappa_n^2 c \Delta} \right)^{3/2}
\ee
Thus, to ensure accuracy in the final sum of Eq. \ref{eq:Jrsigmat}, the bounds of $\sigma$ must be set sufficiently far outside of $\sigma_{tp}$ such that the function is small at the edges. The scale of an $e$-folding in $J_{nm}(\sigma)$ is $k/(\kappa_n \Delta) = \tau_0 / (\sqrt{\pi} n)$, so a grid of $\sigma$ is chosen that spans a large enough number of $e$-foldings that no oscillatory behavior is present at the boundaries of the calculation.

>>>>

We seek now to calculate all eigenmodes $J_{nm}(\sigma)$ and eigenfrequencies $\gamma_{nm}$ for all spatial $n$ and for all eigenmodes $m$ at that given $n$. A resonant response in $J(n,\sigma,-i\gamma)$ is produced for special values of $\gamma \equiv \gamma_{nm}$. To measure the size of the response to detect where resonances occur, we sum the absolute value of $J(n,\sigma,-i\gamma)$ over $\sigma$ in array form. Call this response $f$, and use the index $j$ to represent the value of the response at discrete points over a range of $\gamma$. In places where $f_j > f_{j-1}$ and $f_j>f_{j+1}$, we have bounded a resonance that occurs in the interval $(\gamma_{j-1},\gamma_{j+1})$. We then call the ``solve'' step to hone in on the eigenvalue before continuing the sweep in $\gamma$. In the solve step, we evaluate the eigenfunctions at the points $(\gamma_{j-1},\gamma_{j},\gamma_{j+1})$ to obtain the responses $f_{j-1}$, $f_j$, and $f_{j+1}$. A ``guess'' at the correct eigenvalue $\gamma_{nm}$ can be calculated from
\be
\gamma_{\rm guess} &=& \frac{b\gamma_{j-1} - \gamma_{j+1}}{b - 1}
\ee
where
\be
b &=& \left(\frac{f_{j} - f_{j-1}}{f_{j} - f_{j+1}}\right)\left(\frac{\gamma_{j+1}-\gamma_{j}}{\gamma_{j-1}-\gamma_{j}}\right).
\ee
This is repeated by replacing former points $(\gamma_{j-1},\gamma_{j},\gamma_{j+1})$ with closer estimates while the size of the response grows as the resonance is approached. Given a sufficiently accurate eigenvalue $\gamma_{nm}$, we now find the corresponding eigenfunction $J_{nm}(\sigma)$. We can write down an approximate expression for the response near the resonances as
\be
J(n,\sigma,-i\gamma) & \simeq & \frac{J_{nm}(\sigma)}{\gamma_{nm}-\gamma} + C(\sigma),
\ee
where $C(\sigma)$ is weakly dependent on $\gamma$, and so we treat it as a constant. We can now solve for $J_{nm}(\sigma)$ by comparing two points near the resonance:
\be
J(n,\sigma,-i\gamma_1)  - J(n,\sigma,-i\gamma_2)& = &  \frac{J_{nm}(\sigma)}{\gamma_{nm}-\gamma_1} -  \frac{J_{nm}(\sigma)}{\gamma_{nm}-\gamma_2}
\ee
where $C(\sigma)$ has cancelled out in the difference. We can immediately solve as
\be
J_{nm}(\sigma) & \simeq & \frac{ J(n,\sigma,\gamma_1)  - J(n,\sigma,\gamma_2) }{ 1/(\gamma_{nm}-\gamma_1) - 1/(\gamma_{nm}-\gamma_2)}.
\ee
\begin{figure}
    \centering
    \includegraphics{Jsoln_n1_m100.pdf}
    \caption{Eigenfunctions $J_{nm}(x)$ for the lowest-order spatial eigenmode $n=1$, and a range of frequency eigenmodes up to $m=100$. Their oscillatory forms have varying amplitudes which sum to create the final form of the mean intensity. The largest contribution always comes from the ($n=1, m=1$) lowest-order eigenfunction.}
    \label{fig:jsoln}
\end{figure}
A set of eigenfunctions $J_{nm}(\sigma)$ are plotted in Fig. \ref{fig:jsoln} to illustrate their relative scales for different frequency eigenmodes $m$ at a fixed spatial eigenmode $n$. The overall scale of the $J_{nm}(\sigma)$ are set by the factor $E/(kR^3)$ with $E$ arbitrarily set to 1, $k = \tau_0 \sqrt{\pi} \Delta / R$, and $R = 10^{11}$ cm. With this normalization, an eigenfunction has typical size ${\sim} E a / \left(R^2 \Delta \right) = 10^{-37}$. Additional terms add smaller, faster-oscillating components that lead to higher accuracy upon summation with the lower-order terms. The oscillations of various modes must cancel perfectly to recreate the form of the solution in Eq. \ref{eq:Jrsigmat}, so many modes $m$ and $n$ are required for sufficient numerical precision in the final result. Too few (or imprecisely-calculated) modes creates ``ringing'' in the function that indicates the presence of imperfect cancellations in the positive and negative contributions of individual poles. 
\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{gamma_nm.pdf}
    \caption{The scaling of numerically-obtained resonant frequencies as compared with the analytic expression in Eq. \ref{eq:gamma_nm}.}
    \label{fig:gamma_nm}
\end{figure}
The values of the $\gamma_{nm}$ resonant frequencies are calculable analytically. Their scale depends on $m$, $n$, and the factor $(a\tau_0)^{1/3}$ according to 
\be \label{eq:gamma_nm}
\gamma_{nm} = 2^{-1/3} \pi^{13/6} n^{4/3}\left(m-\frac{7}{8}\right)^{2/3}\frac{c}{R}(a\tau_0)^{-1/3} 
\ee
The size of additional eigenfrequencies scales as a powerlaw in $m$, yet as is seen in Fig. \ref{fig:gamma_nm} the scaling is weak, requiring up to 500 $m$ to reduce the scale of $\gamma_{nm}^{\ \ -1}$ by two orders of magnitude. When sweeping through the imaginary frequency domain to find resonances, this expression is used to set the scale of the sweep points to ensure no $\gamma_{nm}$ are missed. The continuity and close agreement with the analytic expression shown in Fig. \ref{fig:gamma_nm} indicate that the sweep routine accurately brackets and refines the locations of resonances.

The form of a single eigenmode $J_{nm}(\sigma)$ is oscillatory out to some turning point, $\sigma_{tp}$, at which point the function becomes evanescent. The location of the turning point can be found by ignoring the delta-function discontinuity at the source frequency $\sigma_s$ in Eq. \ref{eq:diffusion_plugged_in} and re-examining the resulting homogeneous differential equation. We obtain
\be
\frac{d^2J}{d\sigma^2} & = & \left[ \left( \frac{\kappa_n \Delta }{k} \right)^2 - \frac{3\phi \gamma\Delta^2}{ck}\right] J
\ee
where the line profile can be written approximately as $\phi = (2a/\pi)^{1/3}/(3 \Delta |\sigma|^{2/3})$. When the coefficient on the right hand side is positive, exponential growth or decay solutions are found (evanescence). This occurs on the line wings. When the coefficient on the right hand side is negative, oscillatory solutions are found (propagation), which occurs in the line core. The boundary between propagation and evanescence occurs at the turning point, given by
\be
\sigma_{tp} & = & \sqrt{\frac{2a}{\pi}}\left( \frac{k \gamma}{ \kappa_n^2 c \Delta} \right)^{3/2}
\ee
Thus, to ensure accuracy in the final sum of Eq. \ref{eq:Jrsigmat}, the bounds of $\sigma$ must be set sufficiently far outside of $\sigma_{tp}$ such that the function is small at the edges. The scale of an $e$-folding in $J_{nm}(\sigma)$ is $k/(\kappa_n \Delta) = \tau_0 / (\sqrt{\pi} n)$, so a grid of $\sigma$ is chosen that spans a large enough number of $e$-foldings that no oscillatory behavior is present at the boundaries of the calculation.
